---
title: "Bayesian variable selection for occupancy models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian variable selection for occupancy models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(bvsOccupancy)
```

## What the package does

Occupancy models almost always come with more candidate covariates than the
data can support, and the usual response is to fit a handful of models and
compare them with an information criterion. `bvsOccupancy` instead puts a
binary inclusion indicator on every covariate and samples the model and the
parameters jointly, so the output is a posterior distribution over models
rather than a single selected one.

Three model classes are covered, one function each:

| Function | Model | Components with variable selection |
|---|---|---|
| `bvsSSOM()` | single-season, single-species | occupancy, detection |
| `bvsSDOM()` | dynamic (multi-season), single-species | initial occupancy, persistence, colonisation, detection |
| `bvsMSOM()` | multi-species community | occupancy, detection, per species |

All three share the same machinery. Polya-Gamma data augmentation makes the
regression coefficients conditionally Gaussian, so each block of coefficients
is drawn in one conjugate step. Because the coefficients are conjugate they can
also be integrated out analytically, which is what lets the inclusion vector be
updated with add, delete and swap moves whose acceptance probability depends
only on the marginal likelihood. Categorical covariates are handled as a block:
all dummy columns of a factor enter or leave together, so inference does not
depend on which level happens to be the reference.

## Designs

Every component of a model needs a design object built by `bvsDesign()` for
site-level covariates or `bvsDesignArray()` for covariates that also vary over
visits or seasons. Both take a formula and a data frame, build the model matrix
in the usual R way, and record which columns belong to which covariate.

```{r design, eval = FALSE}
site <- data.frame(
  elev    = rnorm(200),
  forest  = rnorm(200),
  habitat = factor(sample(c("wet", "dry", "mixed"), 200, replace = TRUE))
)

occ <- bvsDesign(~ elev + forest + habitat, site)
occ
```

The printed object shows the covariate groups. `habitat` occupies two columns
but counts as one candidate covariate, and the two columns will always be
selected or dropped together.

Detection covariates vary by visit, so they are supplied as a long data frame
in site-major order, together with the dimensions of the array to build:

```{r design-array, eval = FALSE}
visit <- data.frame(
  date  = rnorm(200 * 5),
  obs   = factor(sample(c("A", "B"), 200 * 5, replace = TRUE))
)
det <- bvsDesignArray(~ date + obs, visit, dims = c(200, 5))
```

The result is a sites by columns by visits array. For a dynamic model the
detection design has a fourth dimension for seasons, `dims = c(M, J, T)`, while
persistence and colonisation designs use `dims = c(M, T - 1)` because they
describe transitions rather than seasons.

Interactions and transformations work exactly as they do in `lm()`. A term such
as `elev:habitat` becomes its own covariate group and is selected as a block.

## Priors

`bvsPrior()` builds the normal prior on the coefficients. The intercept gets a
diffuse variance of its own; the remaining coefficients get a block diagonal
covariance in which a continuous covariate contributes a single variance and
the dummy columns of a factor share an exchangeable block with correlation one
half. That block structure is what makes the fitted effects invariant to the
reference level.

```{r prior, eval = FALSE}
p <- bvsPrior(occ, phi.beta = 1 / 4)
round(p$B, 3)
```

Lowering `phi.beta` shrinks the coefficients harder and favours smaller models.
It is worth refitting at a couple of values to see how sensitive the inclusion
probabilities are, since in variable selection the prior scale on the
coefficients plays the role that a penalty parameter plays elsewhere.

## A single-season model

`simSSOM()` generates data from a known sparse model, which makes it easy to
see whether selection is doing what it should.

```{r ssom, eval = FALSE}
sim <- simSSOM(M = 400, J = 5, n.cont = 3, n.cat = 1, n.levels = 3, seed = 1)

fit <- bvsSSOM(
  y          = sim$y,
  occ.design = sim$occ.design,
  det.design = sim$det.design,
  n.iter     = 15000,
  n.burnin   = 5000,
  n.chains   = 2,
  seed       = 1
)

fit
```

`y` is a sites by visits matrix of zeros and ones, with `NA` for visits that
were not made. Sites may have different numbers of visits; unsurveyed visits
are moved to the end of each row internally, together with their covariates.

The quantity to report is the posterior inclusion probability of each
covariate:

```{r ip, eval = FALSE}
inclusionProbs(fit)
medianModel(fit)
```

The intercept is always in the model, so its probability is one by
construction. `medianModel()` lists the covariates above a threshold, 0.5 by
default, which is the median probability model.

Coefficient summaries come from `summary()`:

```{r summary, eval = FALSE}
summary(fit, pars = "^beta_psi")
```

These need care. Under variable selection the posterior of a coefficient is a
mixture over models that include the covariate and models that do not, and
while a covariate is excluded its coefficient does not enter the likelihood.
A coefficient summary is therefore only interpretable alongside its inclusion
probability. If you want the effect size conditional on inclusion, restrict to
the draws in which the indicator is one:

```{r conditional, eval = FALSE}
draws <- as.matrix(as.mcmc.list(fit))
keep  <- draws[, "gamma_psi[2]"] == 1
quantile(draws[keep, "beta_psi[2]"], c(0.025, 0.5, 0.975))
```

Convergence should be checked as usual, with the caveat that the scale
reduction factor of an indicator that never moves is `NA`, which is expected
for the intercept.

```{r coda, eval = FALSE}
library(coda)
gelman.diag(as.mcmc.list(fit)[, c("beta_psi[2]", "beta_p[2]")])
```

## A dynamic model

For multi-season data, `y` is a sites by visits by seasons array and four
designs are supplied. Persistence and colonisation are defined on transitions,
so their designs cover `T - 1` seasons. If a design with `T` seasons is passed
the last season is dropped with a message.

```{r sdom, eval = FALSE}
sim <- simSDOM(M = 200, J = 5, n.seasons = 10, seed = 1)

fit <- bvsSDOM(
  y          = sim$y,
  occ.design = sim$occ.design,
  per.design = sim$per.design,
  col.design = sim$col.design,
  det.design = sim$det.design,
  n.iter     = 25000,
  n.burnin   = 10000,
  n.chains   = 2,
  seed       = 1
)

inclusionProbs(fit)
```

Selection runs separately on all four components, so a covariate can drive
colonisation without appearing in persistence. If the same covariates are
candidates for both, `col.design` defaults to `per.design` and can be omitted.

Persistence and colonisation are the parameters that are usually weakly
identified in dynamic models, especially with few seasons or low detection. It
is normal for their inclusion probabilities to sit near the prior when the data
carry little information about transitions, and that is a useful finding rather
than a failure.

## A multi-species model

In a community model, `y` is a species by sites by visits array. Species share
community-level means and variances, and selection is species-specific: each
species has its own inclusion vector, drawn towards the community pattern
through the hierarchical prior.

```{r msom, eval = FALSE}
sim <- simMSOM(N = 20, S = 250, nvisits = 5, seed = 1)

fit <- bvsMSOM(
  y          = sim$y,
  occ.design = sim$occ.design,
  det.design = sim$det.design,
  n.iter     = 10000,
  n.burnin   = 5000,
  n.chains   = 2,
  seed       = 1
)

ip <- inclusionProbs(fit, "z")   # species by covariate matrix
round(ip, 2)
colMeans(ip)                     # how community-wide each covariate is
```

`tau.occ` and `tau.det` set the inverse gamma priors on the community
variances, with separate shape and rate for the intercept and for the
covariates. The defaults are moderately informative, which is deliberate: the
community variance is what ties species together, and a very diffuse prior on
it lets species drift apart and weakens selection.

The community covariance uses the same block structure as `bvsPrior()`, with
one variance per covariate group and exchangeable correlation within the dummy
columns of a factor.

## Practical notes

**Compilation.** The model is compiled by nimble the first time it is fitted,
which takes a few minutes and produces C++ compiler output. That cost is paid
once per model structure, not per iteration.

**Run length.** Model indicators mix more slowly than continuous parameters,
because the chain has to move between models rather than within one. The
defaults are a starting point; if inclusion probabilities differ noticeably
between chains, run longer rather than reaching for a different threshold.

**Standardising covariates.** Continuous covariates should be centred and
scaled before they go into a design. The prior covariance is common across
covariates, so unstandardised covariates on wildly different scales receive
effectively different amounts of shrinkage.

**Building on the output.** Setting `run = FALSE` returns the nimble model,
the MCMC configuration and the uncompiled MCMC, so further samplers or monitors
can be added before anything is compiled.

```{r extend, eval = FALSE}
fit <- bvsSSOM(sim$y, sim$occ.design, sim$det.design, run = FALSE)
fit$conf$addMonitors("z")
```

## Reference

The samplers implement the approach described in the accompanying manuscript.
See `citation("bvsOccupancy")` for the current reference.
